Transfer of Learned Opponent Models in Zero Sum Games

### 2021-08-26 Report/paper_draft_2021.Rmd
pass through latest draft. Still a few things to resolve

abstract: |
  Human learning transfer abilities take advantage of important cognitive building blocks such as an abstract representation of concepts underlying tasks and 
  causal models of the environment. One way to build abstract representations of the environment when the task involves interactions with others is to build 
  a model of the opponent that may inform what actions they are likely to take next. In this study, we explore opponent modelling and its transfer with the use of
  computer agents with human-like limited degrees of iterated reasoning. In two experiments, we find that participants
  deviate from Nash equilibrium play and learn to adapt to the opponent's strategy and exploit it. Moreover, we show that participants transfer their learning to new games and that
  this transfer is moderated by the level of sophistication of the opponent. Computational modelling shows that players start each game with a model-based
  learning strategy that facilitates between-game transfer of the opponent's strategy, but then switch to behaviour that is consistent with a model-free learning strategy in the latter stages of the interaction.

@@

A model of the opponent can help increase performance in a particular game, but learning more general characteristics of an opponent may also help increase performance in other games. In this paper, we are specifically interested in the latter: How do people build and use models of their opponent to facilitate learning transfer? Repeated games, in which players interact repeatedly with the same opponent and have the ability to learn about the opponent's strategies and preferences [@mertens1990repeated] are particularly useful to address this question. The early literature on learning transfer in repeated games has mostly focused on the proportion of people who play normatively optimal (e.g. Nash equilibrium play) or use salient actions (e.g. risk dominance) in later games, having had experience with a similar game environment previously [@ho1998iterated; @knez2000]. As is well-known, a Nash equilibrium means that all players in a game play such that no-one can unilaterally improve their performance by deviating from their strategy. When playing against an opponent with a Nash-optimal strategy, I can do no better than play according to the Nash-equilibrium strategy as well. However, when faced with a player who deviates from the Nash-optimal strategy, it may be better for me to exploit this and also deviate from this strategy, increasing my performance beyond what is expected at a Nash equilibrium. Of course, this comes with some risk, as my own deviation from Nash-optimal play may leave myself open to similar exploitation.

Studies that focused on whether people can learn to exploit deviations from Nash-equilibrium play have mostly looked at the ability of players to detect and exploit action-based learning rules [@spiliopoulos2013strategic; @shachat2004we]. These studies used computer opponents that don't adapt play to their human opponent, mostly consisting of playing each action with a fixed probability (a mixed strategy). Findings showed that humans are capable of adapting to non-Nash equilibrium play and detect patterns in opponent's history of play. However, the use of mixed strategies may have limited people's ability to form accurate opponent models. 
Rather than thinking of their opponents and drawing a random action from a (non-uniform) probability distribution, people may rather think of their opponents as applying a form of iterated reasoning to determine their next action. The type of reasoning we refer to takes the form of ``I believe that you believe that I believe ...'' and has been shown to underlie people's decision in the context of economic games [@camerer2003behavioural;@costa2001cognition]. It also underlies successful models in behavioural economics, such as Level-$k$ and Cognitive Hierarchy models [@camerer2004cognitive], which posit that people assume their opponent has a limited level of iterative reasoning, with the player best responding to their (modelled) opponent.

In Level-$k$ theory, a level-0 player uses a fixed strategy without explicitly considering the strategy of their opponent. A level-1 player assumes their opponent is a level-0 player, and chooses actions to best respond to the strategy of their opponent, without considering what their opponent might believe that they will play. A level-2 player, on the other hand, takes their opponent's belief about their actions into account, assuming they face a level-1 player, and choosing actions to best respond to the actions of that player. Cognitive Hierarchy theory works on similar principles, but rather than assuming an opponent always adopts a particular level-$k$ strategy, they assume their opponent adopts each pf the level-$k$ strategies with a particular probability (i.e., the strategy is a mixture over pure level-$k$ strategies).

Iterative reasoning strategies can explain non-equilibrium play in a range of games, such as the p beauty contest game [@nagel1995unraveling], dominance solvable games [@camerer2004cognitive] and various standard normal form games [@costa2001cognition]. In the present study, we endow our computer opponents with limited ability for iterative reasoning and assess whether (1) human players adapt their strategy to exploit this limited reasoning of their opponent, and (2) whether they are able to generalize a learned opponent model to other games. In two experiments, participants face the same opponent (Experiment 1) or same two opponents (Experiment 2) in three consecutive games: the well-known Rock-Paper-Scissors game, a structurally similar Fire-Water-Grass game, and a less similar Numbers (Experiment 1) or Shootout (Experiment 2) game. To foreshadow our results, we find evidence that participants transfer the learned strategy of their opponent to other games, but that this transfer is moderated by the sophistication of the opponent and the similarity of the games. Moreover, using computational modelling, we find evidence that participants switch from relying on an opponent model in the early stages of the games, to a more habiltual and cognitively less demanding strategy in the later stages of the games.

@@

In the first experiment, we aim to test learning transfer through making participants face the same computer opponent in three sequential games that vary in similarity.

@@

### Participants and Design
A total of 52 (28 female, 24 male) participants were recruited on the Prolific Academic platform. The mean age of participants was 31.2 years. Participants were paid a fixed fee of £2.5 plus a bonus dependent on their performance which averaged £1.06. The experiment used a 2 (computer opponent: level 1 or level 2) by 3 (games: rock-paper-scissors, fire-water-grass, numbers) design, with repeated measures on the second factor. Participants were randomly assigned to one of the two levels of the first factor.

@@

Participants played the three games against their computer opponent. These games were rock-paper-scissors, fire-water-grass, and the numbers game. A typical rock-paper-scissors game (hereafter RPS) is a 3x3 zero sum game, with a cyclical hierarchy between the two player's actions: rock blunts scissors, paper wraps rock, and scissors cut paper. If one player chooses an action which dominates their opponent's action, the player wins (receives a reward of 1) and the other player loses (receives a reward of -1). Otherwise it is a draw and both players receive a reward of 0. RPS has a unique mixed-strategy Nash equilibrium, which consists of each player in each round randomly selecting from the three options with uniform probability. The Fire-Water-Grass (FWG) game is identical to RPS in all but action labels: Fire burns grass, water extinguishes fire, and grass absorbs water. We use this game as we are interested in whether learning is transferred in a fundamentally similar game where the only difference is in the name of the possible actions. This should make it relatively easy to generalize knowledge of the opponent's strategy, provided this knowledge is on a sufficiently abstract level, such as knowing the opponent is a level-1 or level-2 player. Crucially, learning simple contingencies such as ``If I played Rock on the previous round, playing Scissors next will likely result in a win'', is not generalizable to this similar game, as these contingencies are tied to the labels of the actions.

@@

The computer opponent was programmed to use either a level-1 or level-2 strategy in all the games. A level-1 player is defined as a player who best responds to a level-0 player. A level-0 player plays in a non-strategic way and does not consider their opponent's actions. Here, we assume a level-0 player simply repeats their previous action. There are other ways to define a level-0 player. For instance, as repeating their action if it resulted in a win and choosing randomly from the remaining actions otherwise, or choosing randomly from all actions. As a best response to a uniformly random action is itself a random action, defining a level-0 player in such a way would make a level-1 opponent's strategy much harder to discern. Because we are mainly interested in generalization of knowledge of an opponent's strategy to other games, which rests on good knowledge of this strategy, we opted for this more deterministic formulation of a level-0 player (whilst also introducing some randomness in the computer opponent's play). A level-2 computer opponent, will assume in turn that the participant is a level-1 opponent, playing according to the strategy just described. We also introduced some noise over the actions of computer opponents making them play randomly in 10\% of all trials. Note that at all levels, the strategies are contingent on the actions taken in the previous round. The choice of this type of strategy is consistent with evidence that humans strategically use information from last round play of their opponents in zero sum games [@batzilis; @wang2014social].

\begin{table}[bth!]
\centering
\begin{tabular}{||c c | c c||} 
 \hline
 Human action $t-1$ & Computer action $t-1$  & Computer level-1 action $t$ & Computer level-2 action $t$ \\ [0.5ex] 
 \hline\hline
 Paper & Rock & Scissors & Scissors \\ 
 Scissors  & Scissors & Rock & Paper \\
 Rock & Paper & Paper & Rock \\
 ... & ... & ... & ... \\ [1ex] 
 \hline
\end{tabular}
\caption{Example of how a level-1 and level-2 computer agent plays in response to actions taken in the previous round.}
\label{table:1}
\end{table}

@@

Participants were informed they would play three different games against the same computer opponent. Participants were told that the opponent cannot cheat and will choose its actions simultaneously with them, without prior knowledge of the participant's choice. After providing informed consent and reading the instructions, participants answered a number of comprehension questions. They then played the three games against their opponent in the order RPS, FGW, and Numbers. An example of the interface for the RPS game is provided in Figure\ \@ref(fig:feedback-rps-exp2). On each round, the human player chooses an action, and after a random delay, is shown the action chosen by the computer opponent, and the outcome of that round. A history of the chosen actions over all rounds is available in the centre panel. A total of 50 rounds of each game was played with the player's score displayed at the end of each game. The score was calculated as the number of wins minus the number of losses. Ties did not affect the score. In order to incentivise the participants to maximise the number of wins against the opponents, players were paid a bonus at the end of the experiment that was proportional to their final score. Each point is worth £0.02. After playing all the games, participants were asked questions about their beliefs about the computer opponent, related to whether they thought they learned their opponent's strategy, and how difficult they found playing against their particular opponent. They were then debriefed and thanked for their participation.

@@

To assess learning within and between games, we used a 2 (condition: Level 1, Level 2) by 3 (game: RPS, FWG, Numbers) by 2 (block: first half, second half) repeated-measures ANOVA, with the first factor varying between participants. This showed a main effect of Game ($F(2,100) = 8.54$, $\eta^{2}  = 0.05$,  $p < .001$), indicating that average scores varied significantly over the games. Post-hoc pairwise comparisons showed that performance in the FWG game was significantly higher than in the RPS game ($t(100) =3.78$, $p < .001$) and the Numbers game ($t(100) = 3.32$ , $p = .002$). The score in RPS was not significantly different from the score in Numbers ($t(100) = 0.45$ , $p = .65$). The main effect of Block ($F(1,50) = 22.51$ ,  $\eta^{2} = 0.03$, $p < .001$) shows that the score in the first half of each game ($M = 0.29$) was significantly lower than in the second half ($M = 0.40$), which indicates within-game learning. The main effect of Condition ($F(1,50) = 5.44$, $\eta^{2} = 0.05$, $p = .024$) indicates that scores were higher against the level-1 player ($M = 0.43$) than against the level-2 player ($M = 0.27$). Thus, it appears that is was harder for participants to exploit the strategy of the more sophisticated level-2 opponent than the comparatively less sophisticated level-1 opponent. **MS: Need to check test results and report GG degrees of freedom, etc.** **...IG: I thought GG corrections only apply when there is violation of sphericity.... according to Mauchly Test there isn't a violation here. Should we report them anyway in this experiment? **

@@

afex::aov_car(
  block_score ~ game.f*condition*block + Error(human_id/(game.f*block))
  , data=dat2
  , type = 3)

@@

As a measure for learning transfer, we focus on participants’ scores in the initial 5 rounds after the first round (rounds 2-6) of each game (see Figure \ref{fig:exp1-early-score-by-opp}). We exclude the very first round as the computer opponent plays randomly here and there is no opportunity yet for the human player to exploit their opponent's strategy. Players with no knowledge of their opponent's strategy are expected to perform at chance level in these early rounds. Positive scores in rounds 2-6 reflect generalization of prior experience. The FWG early score score is significantly higher than 0 ($t(148.85) = 4.584$, $p < .001$). This is also the case for the Numbers game ($t(148.85) = 3.00$, $p = .009$). We did not expect positive scores for the RPS game, as it was the first game played and there was no opportunity for learning about the opponent's strategy. Scores in this game was indeed not significantly different from 0 ($t(148.85) = 1.04$ , $p = .89$).

@@

Next, we explore whether learning transfer is moderated by the type of opponent and game similarity. We expected better transfer between more similar games (i.e. better transfer from RPS to FWG than from RPS/FWG to Numbers), and worse transfer for the more sophisticated level 2 agent. Figure \ref{fig:exp1-early-score-by-opp} indicates that the pattern over the games is indeed dissimilar between level-1 and level-2 opponents. To explore this, we used a 2 (condition: Level 1, Level 2) by 3 (game: RPS, FWG, Numbers) repeated measures ANOVA with the first factor varying between participants. There was a main effect of Game ($F(1.84,91.77) = 3.35$, $\eta^{2}  = 0.04$,  $p = .043$, Greenhouse-Geisser correction to degrees of freedom applied to correct for non-sphericity).  We then run statistical tests on early round scores by game and opponent against the null hypothesis of 0 (no transfer). For level-1 facing players, there is evidence of learning transfer from RPS to both FWG ($t(150) = 3.96$, $p < .001$) and Numbers ($t(150) = 3.74$, $p < .001$). For level-2 facing players, there is evidence for transfer from RPS to the similar game FWG, albeit scores are lower than for level-1 player ($t(150) = 2.48$, $p = .01$) but not to the dissimilar game of Numbers.

@@

In Experiment 2, we aimed to obtain a stronger test of learning transfer. Instead of facing a single level-1 or level-2 opponent throughout all games, participants face both types of opponents sequentially in this experiment. To perform well against both opponents, participants would need to learn distinct strategies in each game. To reduce effects of increased memory load due to facing distinct opponents, we provided participants access to the history of play against an opponent within each game (see Figure \@ref(fig:feedback-rps-exp2)). Finally, we changed the third game to a penalty shootout game, with participants aiming to score a goal and opponents playing the role of goal keepers. Whilst this game has the same number of actions as the first two (aim left, center, or right), it is strategically dissimilar. Unlike the Numbers game in Experiment 1, the Shootout game does not have a cyclical hierarchy between actions, making it harder to win through a heuristic based on this cyclicity.

@@

A total of 48 participants (21 females, 28 males, 1 unknown) were recruited via the Prolific platform, none of which took part in Experiment 1. The average age was 30.2 years, and the mean duration to complete the task was 39 minutes. Participants were incentivised using a two-tier payment mechanism: a fixed fee of £2.5 for completing the experiment plus a performance linked bonus that averaged £1.32.

@@

The three games the participants played were Rock-Paper-Scissors (RPS), Fire-Water-Grass (FWG), and the penalty Shootout game. The first two games were identical to the ones used in the first experiment. In the final game (Shootout) participants took the role of the a football (soccer) player in a penalty situation, with the computer opponent taking the role of the goalkeeper. Players had the choice between three actions: shooting the football to the left, right or centre of the goal. Similarly, the goalkeeper chooses between defending the left, right, or centre of the goal. If participants shoot in a different direction than where the goalkeeper defends, they win the round and the goalkeeper loses. Otherwise, the goalkeeper catches the ball and the player loses the round. There is no possibility of ties in this game. Figure\ \@ref(fig:screenshot-shootout) shows a snapshot of play in the shootout game. What makes this game different to the other games is that there are two ways to beat the opponent: if the shooter thinks their opponent is going to choose to defend ''right'' in the next round, they can win by either choosing to shoot ''left'' or ''center''. A level-1 shooter who thinks that their goalkeeper opponent will repeat their last action has thus two possible best responses. A level-1 goalkeeper, however, has only a single best response (defending where their opponent aimed in the last round). A level-2 goalkeeper, who believes their opponent is a level-1 shooter, will have two best responses however. We programmed the level-2 computer player to choose randomly between these two best responses.

@@

Participants played 3 games sequentially against both level-1 and level-2 computer opponents, rather than just one like in the first experiment. Like in the first experiment, the computer opponents retained the same strategy throughout the 3 games, however the participants faced each opponent twice in each game. Specifically, each game was divided into 4 stages numbered 1 to 4, consisting of 20, 20, 10, and 10 rounds respectively for a total of 60 rounds per game. Participants start by facing one of the opponents in stage one, then face the other in stage two. This is repeated in the same order in stages 3 and 4. Which opponent they faced first was counterbalanced. All participants engage in the same three games (namely RPS, FWG and Shootout) in this exact order, and were aware that the opponent was not able to know their choices beforehand but was choosing simultaneously with the player. 
In order to encourage participants to think about their next choice, a countdown timer of 3 seconds was introduced at the beginning of each round. During those 3 seconds, participants could not choose an action and had to wait for the timer to run out. A small random delay between 0.5 and 3 seconds was also introduced before the computer agent made their choice, as a way of simulating a real human opponent thinking time. After each round, participants were given detailed feedback about their opponent's action and whether they won or lost the round. Further information about the history of play in previous rounds was also provided and participants could scroll down to recall the full history of each interaction against an opponent in a particular stage of a game. The number of wins, losses and ties were clearly shown at the top of the screen for each game, and this scoreboard was reinitialised to zero at the onset of a new stage game.
As in the first experiment, all the games have a unique mixed-strategy Nash equilibrium consisting of uniformly random actions. If participants follow this strategy, or simply don't engage in learning how the opponent plays, they would score 0 on average against both level-1 and level-2 players. Evidence of sustained wins would indicate that participants have learned to exploit patterns in their opponents' play.

@@

Participants scores are depicted in Figure \ref{fig:exp2-score-by-opp}. The RPS game had the lowest average score per round ($M = 0.194$, SD = 0.345) followed by FWG ($M = 0.27$, SD = 0.394) and finally the Shootout game ($M = 0.289$, SD = 0.326).^[ A higher score in shootout is expected as there are 2 out of three possible winning actions, compared to one out of three in RPS and FWG. Indeed, a player not aiming to uncover the opponent’s strategy and thus choosing to play randomly should be expected to have on average score per round of 0 in both RPS and FWG, and 0.33 in the Shootout game. To make the scores more comparable, and because we are interested in player’s performance that is not due to chance, we will adjust all scores in the shootout game by subtracting the average score per round of a random strategy (0.33)] Using parametric t-tests on adjusted scores, we reject the null hypothesis of random play in all three games (RPS: $t(49) = 6.26$, $p < .001$; FWG: $t(49) = 7.25$, $p < .001$; Shootout: $t(49) = 13.61$, $p < .001$).

@@

data_long_post1 <- gather(mean_post_1, strategy, probability, Nash_posterior, RL_posterior,BCH_BT_posterior, factor_key=TRUE)

>>>>

abstract: |
  Human learning transfer abilities take advantage of important cognitive building blocks such as an abstract representation of concepts underlying tasks and 
  causal models of the environment. One way to build abstract representations of the environment when the task involves interactions with others is to build 
  a model of the opponent that may inform what actions they are likely to take next. In this study, we explore opponent modelling and its transfer with the use of
  computer agents with human-like limited degrees of iterated reasoning. In two experiments, we find that participants
  deviate from Nash equilibrium play and learn to adapt to their opponent's strategy to exploit it. Moreover, we show that participants transfer their learning to new games and that
  this transfer is moderated by the level of sophistication of the opponent, as well as the similarity between games. Computational modelling shows that players start each game with a model-based
  learning strategy that facilitates between-game transfer of their opponent's strategy, but then switch to behaviour that is consistent with a model-free learning strategy in the latter stages of the interaction.

@@

A model of the opponent can help increase performance in a particular game, but learning more general characteristics of an opponent may also help increase performance in other games. In this paper, we are specifically interested in the latter: How do people build and use models of their opponent to facilitate learning transfer? Repeated games, in which players interact repeatedly with the same opponent and have the ability to learn about their opponent's strategies and preferences [@mertens1990repeated] are particularly useful to address this question. The early literature on learning transfer in repeated games has mostly focused on the proportion of people who play normatively optimal (e.g. Nash equilibrium play) or use salient actions (e.g. risk dominance) in later games, having had experience with a similar game environment previously [@ho1998iterated; @knez2000]. As is well-known, a Nash equilibrium means that all players in a game act such that no-one can unilaterally improve their performance by deviating from their strategy. When playing against an opponent with a Nash-optimal strategy, I can do no better than play according to the Nash-equilibrium strategy as well. However, when faced with a player who deviates from the Nash-optimal strategy, I may be able to exploit this and also deviate from this strategy, increasing my performance beyond what is expected at Nash equilibrium. Of course, this comes with some risk, as my own deviation from Nash-optimal play may leave me open to similar exploitation.

Studies that focused on whether people can learn to exploit deviations from Nash-equilibrium play have mostly looked at the ability of players to detect and exploit action-based learning rules [@spiliopoulos2013strategic; @shachat2004we]. These studies used computer opponents that don't adapt play to their human opponent, mostly consisting of playing each action with a fixed probability (a mixed strategy). Findings showed that humans are capable of adapting to non-Nash-equilibrium play and detect patterns in an opponent's history of play. However, the use of mixed strategies may have limited people's ability to form accurate opponent models. 
Instead of thinking of their opponents as drawing random actions from a (non-uniform) probability distribution, people may rather think of their opponents as applying a form of iterated reasoning to determine their next action. The type of reasoning we refer to takes the form of "I believe that you believe that I believe ..." and has been shown to underlie people's decisions in economic games [@camerer2003behavioural;@costa2001cognition]. It also underlies successful models in behavioural economics, such as Level-$k$ and Cognitive Hierarchy models [@camerer2004cognitive], which posit that people assume their opponent applies a limited level of iterative reasoning, taking actions that are best responses to those of their (modelled) opponent.

In Level-$k$ theory, a level-0 player uses a fixed strategy without explicitly considering the strategy of their opponent. A level-1 player assumes their opponent is a level-0 player, and chooses actions to best respond to the strategy of their opponent, without considering what their opponent might believe that they will play. A level-2 player, on the other hand, takes their opponent's belief about their actions into account, assuming they face a level-1 player, and choosing actions to best respond to the actions of that player. Cognitive Hierarchy theory is based on similar principles, but rather than assuming an opponent always adopts a particular level-$k$ strategy, they are assumed to adopt each of the level-$k$ strategies with a particular probability (i.e., the opponent's strategy is a mixture over pure level-$k$ strategies).

Iterative reasoning strategies can explain non-equilibrium play in a range of games, such as the $p$-beauty contest game [@nagel1995unraveling], dominance solvable games [@camerer2004cognitive], and various standard normal form games [@costa2001cognition]. In the present study, we endow our computer opponents with limited ability for iterative reasoning and assess whether (1) human players adapt their strategy to exploit this limited reasoning of their opponent, and (2) whether they are able to generalize a learned opponent model to other games. In two experiments, participants face the same opponent (Experiment 1) or the same two opponents (Experiment 2) in three consecutive games: the well-known Rock-Paper-Scissors game, the structurally similar Fire-Water-Grass game, and a less similar Numbers (Experiment 1) or Shootout (Experiment 2) game. To foreshadow our results, we find evidence that participants transfer the learned strategy of their opponent to other games, but that this transfer is moderated by the sophistication of their opponent and the similarity of the games. Moreover, using computational modelling, we find evidence that participants switch from relying on an opponent model in the early stages of the games, to a more habitual and cognitively less demanding strategy in the later stages of the games.

@@

In the first experiment, we aim to test learning transfer by making participants face the same computer opponent with a limited level of iterative reasoning in three sequential games that vary in similarity. If participants are able to learn the limitations of their opponent's iterative reasoning and generalize this to new games, their performance in (the early stages of) later games should be higher than expected if they were to completely learn a new strategy in each game.

@@

### Participants and Design

A total of 52 (28 female, 24 male) participants were recruited on the Prolific Academic platform. The mean age of participants was 31.2 years. Participants were paid a fixed fee of £2.5 plus a bonus dependent on their performance (£1.06 on average). The experiment had a 2 (computer opponent: level 1 or level 2) by 3 (games: rock-paper-scissors, fire-water-grass, numbers) design, with repeated measures on the second factor. Participants were randomly assigned to one of the two levels of the first factor.

@@

Participants played the three games against their computer opponent. These games were Rock-Paper-Scissors, Fire-Water-Grass, and the Numbers game. A typical Rock-Paper-Scissors game (hereafter RPS) is a 3x3 zero-sum game, with a cyclical hierarchy between the two player's actions: rock blunts scissors, paper wraps rock, and scissors cut paper. If one player chooses an action which dominates their opponent's action, the player wins (receives a reward of 1) and the other player loses (receives a reward of -1). Otherwise it is a draw and both players receive a reward of 0. RPS has a unique mixed-strategy Nash equilibrium, which consists of each player in each round randomly selecting from the three options with uniform probability.

The Fire-Water-Grass (FWG) game is identical to RPS in all but action labels: Fire burns grass, water extinguishes fire, and grass absorbs water. We use this game as we are interested in whether learning is transferred in a fundamentally similar game where the only difference is in the label of the possible actions. This should make it relatively easy to generalize knowledge of the opponent's strategy, provided this knowledge is on a sufficiently abstract level, such as knowing the opponent is a level-1 or level-2 player. Crucially, learning simple contingencies such as "If I played rock on the previous round, playing scissors next will likely result in a win", is not generalizable to this similar game, as these contingencies are tied to the labels of the actions.

@@

The computer opponent was programmed to use either a level-1 or level-2 strategy in all the games. A level-1 player is defined as a player who best responds to a level-0 player. A level-0 player plays in a non-strategic way and does not consider their opponent's actions. Here, we assume a level-0 player simply repeats their previous action. There are other ways to define a level-0 player. For instance, as repeating their action if it resulted in a win and choosing randomly from the remaining actions otherwise, or choosing randomly from all actions. As a best response to a uniformly random action is itself a random action, defining a level-0 player in such a way would make a level-1 opponent's strategy much harder to discern. Because we are mainly interested in generalization of knowledge of an opponent's strategy to other games, which requires good knowledge of this strategy, we opted for this more deterministic formulation of a level-0 player. The level-2 computer opponent assumes in turn that the participant is a level-1 opponent, playing according to the strategy just described. To make the computer opponent's strategy not too obvious, we introduced some randomness in actions of computer opponent, making them play randomly in 10\% of all trials. Note that at all levels, the strategies are contingent on the actions taken in the previous round. The choice of this type of strategy is consistent with evidence that humans strategically use information from last round play of their opponents in zero-sum games [@batzilis; @wang2014social]. Table \@ref(table:1) shows an example of the computer opponent's actions in response to the previous round play.

\begin{table}[bth!]
\caption{Example of how a level-1 and level-2 computer agent plays in response to actions taken in the previous round.}
\centering
\label{table:1}
\begin{tabular}{||c c | c c||} 
 \hline
 Human action $t-1$ & Computer action $t-1$  & Computer level-1 action $t$ & Computer level-2 action $t$ \\ [0.5ex] 
 \hline\hline
 Paper & Rock & Scissors & Scissors \\ 
 Scissors  & Scissors & Rock & Paper \\
 Rock & Paper & Paper & Rock \\
 ... & ... & ... & ... \\ [1ex] 
 \hline
\end{tabular}
\end{table}

@@

Participants were informed they would play three different games against the same computer opponent. Participants were told that the opponent cannot cheat and will choose its actions simultaneously with them, without prior knowledge of the participant's choice. After providing informed consent and reading the instructions, participants answered a number of comprehension questions. They then played the three games against their opponent in the order RPS, FGW, and Numbers. An example of the interface for the RPS game is provided in Figure\ \@ref(fig:feedback-rps-exp2). On each round, the human player chooses an action, and after a random delay (between 0.5 and 3 seconds) is shown the action chosen by the computer opponent and the outcome of that round. <!-- A history of the chosen actions over all rounds is available in the centre panel. **MS: was this not only the case for Experiment 2? ** --> A total of 50 rounds of each game was played with the player's score displayed at the end of each game. The score was calculated as the number of wins minus the number of losses. Ties did not affect the score. In order to incentivise the participants to maximise the number of wins against their opponent, players were paid a bonus at the end of the experiment proportional to their final score (each point was worth £0.02). After playing all the games, participants were asked questions about their beliefs about the computer opponent, related to whether they thought they learned their opponent's strategy, and how difficult they found playing against their opponent. They were then debriefed and thanked for their participation.

@@

To assess learning within and between games, we used a 2 (condition: Level 1, Level 2) by 3 (game: RPS, FWG, Numbers) by 2 (block: first half, second half) repeated-measures ANOVA, with the first factor varying between participants. This showed a main effect of Game ($F(2,100) = 8.54$, $\eta^{2}  = 0.05$,  $p < .001$), indicating that average scores varied significantly over the games. Post-hoc pairwise comparisons showed that performance in the FWG game was significantly higher than in the RPS game ($t(100) =3.78$, $p < .001$) and the Numbers game ($t(100) = 3.32$ , $p = .002$). The score in RPS was not significantly different from the score in Numbers ($t(100) = 0.45$ , $p = .65$). The main effect of Block ($F(1,50) = 22.51$ ,  $\eta^{2} = 0.03$, $p < .001$) shows that the score in the first half of each game ($M = 0.29$) was significantly lower than in the second half ($M = 0.40$), which indicates within-game learning. The main effect of Condition ($F(1,50) = 5.44$, $\eta^{2} = 0.05$, $p = .024$) indicates that scores were higher against the level-1 player ($M = 0.43$) than against the level-2 player ($M = 0.27$). Thus, it appears that is was harder for participants to exploit the strategy of the more sophisticated level-2 opponent than the comparatively less sophisticated level-1 opponent.

@@

tmp <- afex::aov_car(
  block_score ~ game.f*condition*block + Error(human_id/(game.f*block))
  , data=dat2
  , type = 3)

@@

As a measure for learning transfer, we focus on participants’ scores in the initial 5 rounds after the first round (rounds 2-6) of each game (see Figure \ref{fig:exp1-early-score-by-opp}). We exclude the very first round as the computer opponent plays randomly here and there is no opportunity yet for the human player to exploit their opponent's strategy. Players with no knowledge of their opponent's strategy are expected to perform at chance level in these early rounds. Positive scores in rounds 2-6 reflect generalization of prior experience. The FWG early score score is significantly higher than 0 ($t(148.85) = 4.584$, $p < .001$). This is also the case for the Numbers game ($t(148.85) = 3.00$, $p = .009$). We did not expect positive scores for the RPS game, as it was the first game played and there was no opportunity for learning about the opponent's strategy. Scores in this game were indeed not significantly different from 0 ($t(148.85) = 1.04$ , $p = .89$).

@@

Next, we explore whether learning transfer is moderated by the type of opponent and game similarity. We expected better transfer between more similar games (i.e. better transfer from RPS to FWG than from RPS/FWG to Numbers), and worse transfer for the more sophisticated level-2 agent. Figure \ref{fig:exp1-early-score-by-opp} indicates that the pattern over the games is indeed dissimilar between level-1 and level-2 opponents. To explore this, we used a 2 (Condition: Level 1, Level 2) by 3 (Game: RPS, FWG, Numbers) repeated measures ANOVA with the first factor varying between participants. There was a main effect of Game ($F(1.84,91.77) = 3.35$, $\eta^{2}  = 0.04$,  $p = .043$, Greenhouse-Geisser correction applied to the degrees of freedom to correct for non-sphericity). **MS: was this the only significant effect? If so, people might wonder why we next run analyses for each combination of player and game** We then ran statistical tests on early round scores by game and opponent, against the null hypothesis of 0 (no transfer).  For level-1 facing players, there is evidence of learning transfer from RPS to both FWG ($t(150) = 3.96$, $p < .001$) and Numbers ($t(150) = 3.74$, $p < .001$). For level-2 facing players, there is evidence for transfer from RPS to the similar game FWG, albeit scores are lower than for level-1 player ($t(150) = 2.48$, $p = .01$) but not to the dissimilar game of Numbers.

@@

In Experiment 2, we aimed to obtain a stronger test of learning transfer. Instead of facing a single level-1 or level-2 opponent throughout all games, participants now faced both types of opponent. To perform well against both opponents, participants would need to learn distinct strategies against these opponents. To reduce effects of increased memory load due to facing distinct opponents, we provided participants access to the history of play against an opponent within each game (see Figure \@ref(fig:feedback-rps-exp2)). Finally, we changed the third game to a penalty Shootout game, with participants aiming to score a goal and opponents playing the role of goal keepers. Whilst this game has the same number of actions as the first two (aim left, center, or right), it is strategically dissimilar. Unlike the Numbers game in Experiment 1, the Shootout game does not have a cyclical hierarchy between actions, making it harder to win through a heuristic based on this cyclicity.

@@

A total of 48 participants (21 females, 28 males, 1 unknown) were recruited via the Prolific platform, none of which took part in Experiment 1. The average age was 30.2 years, and the mean duration to complete the task was 39 minutes. Participants received a fixed fee of £2.5 for completing the experiment and a performance dependent bonus (£1.32 on average).

@@

Participants played three games: Rock-Paper-Scissors (RPS), Fire-Water-Grass (FWG), and the penalty Shootout game. The first two games were identical to the ones used in Experiment 1. In the Shootout game, participants took the role of the a football (soccer) player in a penalty situation, with the computer opponent taking the role of the goalkeeper. Players had the choice between three actions: shooting the football to the left, right, or centre of the goal. Similarly, the goalkeeper chooses between defending the left, right, or centre of the goal. If participants shoot in a different direction than where the goalkeeper defends, they win the round and the goalkeeper loses. Otherwise, the goalkeeper catches the ball and the player loses the round. There is no possibility of ties in this game. Figure\ \@ref(fig:screenshot-shootout) shows a snapshot of play in the Shootout game. What makes this game different to the other games is that there are now two ways to beat the opponent: if the shooter thinks their opponent is going to choose to defend "right" in the next round, they can win by either choosing to shoot "left" or "center". A level-1 shooter who thinks that their goalkeeper opponent will repeat their last action has thus two possible best responses. A level-1 goalkeeper, however, has only a single best response (defending where their opponent aimed in the last round). A level-2 goalkeeper, who believes their opponent is a level-1 shooter, will have two best responses however. We programmed the level-2 computer player to choose randomly between these two best responses.

As in Experiment 1, all the games have a unique mixed-strategy Nash equilibrium consisting of uniformly random actions. If participants follow this strategy, or simply don't engage in learning how the opponent plays, they would score 0 on average against both level-1 and level-2 players. Evidence of sustained wins would indicate that participants have learned to exploit patterns in their opponents' play.

@@

Participants played 3 games sequentially against both level-1 and level-2 computer opponents. As in Experiment 1, the computer opponents retained the same strategy throughout the 3 games. Participants faced each opponent twice in each game. Each game was divided into 4 stages, numbered 1 to 4, consisting of 20, 20, 10, and 10 rounds respectively for a total of 60 rounds per game. Participants started by facing one of the opponents in stage one, then the other in stage two. This was repeated in the same order in stages 3 and 4. Which opponent they faced first was counterbalanced. All participants engaged in the three games (RPS, FWG and Shootout) in this exact order, and were aware that their opponent could not cheat and chose their action simultaneously with the player, without knowing their choices beforehand. In order to encourage participants to think about their next choice, a countdown timer of 3 seconds was introduced at the beginning of each round. During those 3 seconds, participants could not choose an action and had to wait for the timer to run out. A random delay between 0.5 and 3 seconds was again introduced before the choice of the computer agent was revealed, as a way of simulating a real human opponent's decision time. After each round, participants were given detailed feedback about their opponent's action and whether they won or lost the round. Further information about the history of play in previous rounds was also provided and participants could scroll down to recall the full history of each interaction against an opponent in a particular stage of a game. The number of wins, losses and ties in each game were displayed at the top of the screen, and this scoreboard was reinitialised to zero at the onset of a new stage game.

@@

Participants' scores are depicted in Figure \ref{fig:exp2-score-by-opp}. The RPS game had the lowest average score per round ($M = 0.194$, SD = 0.345) followed by FWG ($M = 0.27$, SD = 0.394) and finally the Shootout game ($M = 0.289$, SD = 0.326).^[A higher score in the Shootout game is expected as there are 2 out of three possible winning actions, compared to one out of three in RPS and FWG. Indeed, a player not aiming to uncover the opponent’s strategy and thus choosing to play randomly should be expected to have on average score per round of 0 in both RPS and FWG, and 0.33 in the Shootout game. To make the scores more comparable, and because we are interested in player’s performance that is not due to chance, we will adjust all scores in the Shootout game by subtracting the average score per round of a random strategy (0.33).] Using one-sample t-tests on adjusted scores, we reject the null hypothesis of random play in all three games (RPS: $t(49) = 6.26$, $p < .001$; FWG: $t(49) = 7.25$, $p < .001$; Shootout: $t(49) = 13.61$, $p < .001$).

@@

data_long_post1 <- gather(mean_post_1, strategy, probability, Nash, RL,BCH_BT, factor_key=TRUE)
